---
title: "class06"
author: "Maddie Maslyar (PID: A69042845)"
format: pdf
---

# R Functions Lab

I am submitting the grade function R script instead of answering the "Generate Random Protein In Fasta Format" prompt because I was absent from class.

Q1. Write a function grade() to determine an overall grade from a vector of student homework assignment scores dropping the lowest single score.

```{r}

# Function grade() determines grade from a vector of student homework assignment 
#scores dropping the lowest score.
grade <- function(x) {
  # Set missing homework values (NA) to zero
  x[is.na(x)] <- 0
  # Exclude the lowest homework score
  mean(x[-which.min(x)])
}

```



```{r}
# Input example gradebook in csv format to test grade().
url <- "https://tinyurl.com/gradeinput"
gradebook <- read.csv(url, row.names=1)
```

Q2. Who is the top scoring student in the gradebook?

```{r}
results <- apply(gradebook, 1, grade)
sort(results, decreasing = T)
```

**Student 18 is the top scoring student.**

Q3. Determine the most difficult homework assignment.

```{r}
hw.mean <- apply(gradebook, 2, mean, na.rm = T)
hw.median <- apply(gradebook, 2, median, na.rm = T)
sort(hw.mean)
sort(hw.median)
```

**HW3 and HW2 were the most difficult assignments. HW3 has the lowest mean and HW2 has the lowest median.**
